Liveness-Based Garbage Collection

### 2014-07-23 paper/sections/intro.tex
intro new

Most modern  programming languages support dynamic  allocation of heap
data.  Static  analysis of heap data  is much harder  than analysis of
static   and   stack   data.    Garbage   collectors,   for   example,
conservatively  approximate  the liveness  of  heap  objects by  their
reachability  from a  set of  memory  locations called  the {\em  root
  set\/}.  Consequently, many objects  that are reachable but not live
remain  uncollected, causing  a  larger-than-necessary memory  demand.
This is confirmed  by empirical studies on Haskell~\cite{rojemo96lag},
Scheme~\cite{karkare06effectiveness}                                and
Java~\cite{shaham02estimating} programs.

Here we consider a first-order  pure functional language and propose a
liveness  analysis  which  annotates  various program  points  with  a
description  of variables and  fields whose  object references  may be
dereferenced  in the future.   The garbage  collector then  only marks
objects pointed by live references and leaves other, merely reachable,
objects  to  be  reclaimed.    (Although  not  strictly  necessary,  a
collector  would  normally {\em  nullify}  dead  variables and  fields
rather than leaving dangling  references.)  Since there are fewer live
objects   than   reachable   objects,   more  memory   is   reclaimed.
Additionally, since  the collector traverses a smaller  portion of the
heap, the time spent for each collection is also smaller.  The work is
presented in  the context of a  stop-the-world non-incremental garbage
collector (mark-and-sweep,  compacting or  copying) for which  we also
show a  monotonicity result: that  our technique can never  cause more
garbage  collections  to  occur   in  spite  of  changing  the  rather
unpredictable  execution  points   at  which  collections  occur.   We
anticipate that our technique  is applicable to more modern collectors
(generational,  concurrent, parallel),  but leave  such  extensions to
future work.

We first define a {\em fully context-sensitive} (in the sense that its
results  are unaffected  by function  inlining) liveness  analysis and
prove it  correct.  However, fully context-sensitive  methods often do
not scale,  and this analysis would  also require us  to determine, at
run-time,  the internal  liveness of  a  function body  at each  call.
Hence,   similarly   to   the   0-CFA   approach,   we   determine   a
context-independent {\em summary} of  liveness for each function which
safely   approximates   the   context-dependence   of   all   possible
calls~\cite{chatterjee99relevant,cherem07fast,lee05static}.       (Note
that  an intraprocedural context-insensitive  method which  assumes no
information  about function  callers would  be too  imprecise  for our
needs.)   In essence  our approach  sets up  interprocedural data-flow
equations for the liveness summaries  of functions and shows how these
can be solved symbolically  as context-free grammars (CFGs)\@.  We can
then determine  a CFG  for each program  point; these are  then safely
approximated with  finite-state automata  which are encoded  as tables
for each program point.  For  garbage collection purposes only {\em GC
  points}  (program points  representing  calls, to  $\CONS$  or to  a
user-function) need to be stored.
% which may call eventually call $\CONS$, need to be stored.

We  previously proposed  an intraprocedural  method for  heap liveness
analysis   for   a   Java-like   language~\cite{khedker07heap}   which
statically inserted  statements nullifying dead  references to improve
garbage collection; by  contrast nullification here occurs dynamically
(which can work better with  aliasing) when the garbage collector acts
on  liveness  annotations  to  avoid traversing  dead  references.   A
workshop    paper~\cite{karkare07liveness}    outlined    the    basic
0-CFA-style-summary  interprocedural  approach  to  functional-program
liveness  analysis.   The  current  paper adds  the  context-sensitive
analysis and better formalization along with experimental results.
%   to  substantiate our ideas.

\kern -3pt   %% AM: a very small tweak to get back to 2 pages for this
             %% section.
%% AM: rewrite the next paragraph to save 3-4 words and remove this \kern.

%------------------------------------------------------------%
\setcounter{page}{2}
\subsubsection{Motivating Example}
\label{sec:motiv}

>>>>

Functional  programming  languages  depend  on  garbage  collector  to
efficiently  reclaim  unused  memory   allocated  by  programs  either
explicitly (for example using constructors) or implicitly (for example
for      closures).       However,      empirical      studies      on
Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
Java~\cite{shaham02estimating}  programs   have  shown   that  garbage
collectors leave uncollected a lot of memory objects that are not live
(here {\em live}ness  means the object can potentially be  used by the
program at  a later  stage).  Due  to lack of  a suitable  analysis to
distinguish  reachable  objects  from live  objects.  current  garbage
collectors conservatively approximate the  liveness of heap objects by
their reachability from a set of memory locations called the {\em root
  set\/} resulting in unnecessary memory usage.

Languages having lazy  semantics for evaluation put  further demand on
memory as they  require heap allocated closure to be  carried from the
point of occurrence  of an expression to the point  of (first) real use
of  that expression.  Strictness analysis~\cite{some-refs}  statically
determines  the  patterns  of  future usage  of  an  expression,  thus
allowing some  of the  closures to  be evaluated at  the point  of the
occurrence of the corresponding expression.

We propose the  use of liveness analysis of heap  cells and strictness
analysis  of expressions  to  reduce the  memory  requirement of  lazy
functional  programming languages.   We modify  garbage collectors  to
preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
This results in an increase  in the garbage reclaimed and consequently
in fewer  garbage collections.  Further, we eagerly evaluate closures
that are guaranteed to be evaluated in the future (strict closures).

The main contribution  of this paper is an uniform  framework for {\em
  liveness} and {\em strictness}  analysis to enable the optimizations
mentioned  above. Using  a central  notion of  demand, we  formulate a
context-sensitive liveness  analysis for structured data  and prove it
correct.  We then  present strictness as a dual of  liveness. We use a
0-CFA-like conservative approximation to  annotate each allocation and
function-call program point with  a finite-state automaton---which the
garbage-collector inspects to curtail reachability during marking. The
result of  the strictness analysis,  being necessarily finite,  can be
directly used to insert code to evaluate closures.

We previously proposed an interprocedural liveness analysis to improve
garbage collection first order eager subset of Scheme~\cite{cc-paper}.
The current paper extends the work to handle lazy evaluation semantics
for the same language.

Experiments  confirm  the  expected  performance benefits.  The  early
evaluation   of   closures  decreases   the   memory   demands  of   a
program. Further,  liveness-based collection results in  a increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections, a decrease  in the memory size required  to run programs,
and  reduced  overall  garbage  collection  time  for  a  majority  of
programs.

%------------------------------------------------------------%
\subsubsection{Motivating Example}
\label{sec:motiv}
